\chapter{Network Protocol}\label{chap:net}

The SANE interface has been designed to facilitate network access to
image acquisition devices.  In particular, most SANE implementations
are expected to support a network backend (net client) and a
corresponding network daemon (net server) that allows accessing image
acquisition devices through a network connection.  Network access is
useful in several situations:
\begin{itemize}

\item To provide controlled access to resources that are inaccessible
  to a regular user.  For example, a user may want to access a device
  on a host where she has no account on.  With the network protocol,
  it is possible to allow certain users to access scanners without
  giving them full access to the system.

  Controlling access through the network daemon can be useful even in
  the local case: for example, certain backends may require root
  privileges to access a device.  Rather than installing each frontend
  as setuid-root, a system administrator could instead install the
  SANE network daemon as setuid-root.  This enables regular users to
  access the privileged device through the SANE daemon (which,
  presumably, supports a more fine-grained access control mechanism
  than the simple setuid approach).  This has the added benefit that
  the system administrator only needs to trust the SANE daemon, not
  each and every frontend that may need access to the privileged
  device.

\item Network access provides a sense of ubiquity of the available
  image acquisition devices.  For example, in a local area network
  environment, this allows a user to log onto any machine and have
  convenient access to any resource available to any machine on the
  network (subject to permission constraints).

\item For devices that do not require physical access when used (e.g.,
  video cameras), network access allows a user to control and use
  these devices without being in physical proximity.  Indeed, if such
  devices are connected to the Internet, access from any place in the
  world is possible.

\end{itemize}

The network protocol described in this chapter has been design with
the following goals in mind:
\begin{enumerate}

\item Image transmission should be efficient (have low encoding
  overhead).

\item Accessing option descriptors on the client side must be
  efficient (since this is a very common operation).

\item Other operations, such as setting or inquiring the value of an
  option are less performance critical since they typically require
  explicit user action.

\item The network protocol should be simple and easy to implement on
  any host architecture and any programming language.

\end{enumerate}
The SANE protocol can be run across any transport protocol that
provides reliable data delivery.  While SANE does not specify a
specific transport protocol, it is expected that TCP/IP will be among
the most commonly used protocols.

\section{Data Type Encoding}

\subsection{Primitive Data Types}

The four primitive types of the SANE standard are encoded as follows:
\begin{description}

\item[\code{\defn{SANE\_Byte}}:] A byte is encoded as an 8 bit value.
  Since the transport protocol is assumed to be byte-orientd, the bit
  order is irrelevant.

\item[\code{\defn{SANE\_Word}}:] A word is encoded as 4 bytes (32
  bits).  The bytes are ordered from most-significant to
  least-significant byte (big-endian byte-order).

\item[\code{\defn{SANE\_Char}}:] A character is currently encoded as an 8-bit
  ISO LATIN-1 value.  An extension to support wider character sets (16 or 32
  bits) is planned for the future, but not supported at this point.

\item[\code{\defn{SANE\_String}}:] A string pointer is encoded as a
  \code{SANE\_Char} array.  The trailing NUL byte is considered part
  of the array and a \code{NULL} pointer is encoded as a zero-length
  array.
  
\item[\code{\defn{SANE\_Handle}}:] A handle is encoded like a word.
  The network backend needs to take care of converting these integer
  values to the opaque pointer values that are presented to the user
  of the network backend.  Similarly, the SANE daemon needs to take
  care of converting the opaque pointer values it receives from its
  backends into 32-bit integers suitable for use for network encoding.

\item[{\em\defn{enumeration types}}:] Enumeration types are encoded
  like words.

\end{description}

\subsection{Type Constructors}

Closely following the type constructors of the C language, the SANE network
protocol supports the following four constructors:
\begin{description}

\item[{\em\defn{pointer}}:] A pointer is encoded by a word that indicates
  whether the pointer is a NULL-pointer which is then followed by the
  value that the pointer points to (in the case of a non-NULL pointer;
  in the case of a NULL pointer, no bytes are encoded for the pointer
  value).

\item[{\em\defn{array}}:] An array is encoded by a word that indicates
  the length of the array followed by the values of the elements in
  the array.  The length may be zero in which case no bytes are
  encoded for the element values.

\item[{\em\defn{structure}}:] A structure is encoded by simply encoding the
  structure members in the order in which they appear in the
  corresponding C type declaration.

\item[{\em\defn{union}}:] A union must always be accompanied by a tag
  value that indicates which of the union members is the currently the
  active one.  For this reason, the union itself is encoded simply by
  encoding the value of the currently active member.

\end{description}

Note that for type constructors, the pointer, element, or member
values themselves may have a constructed type.  Thus, the above rules
should be applied recursively until a sequence of primitive types has
been found.

Also SANE had no need for encoding of circular structures.  This
greatly simplifies the network protocol.

\section{Remote Procedure Call Requests}

The SANE network protocol is a client/server-style remote procedure
call (RPC) protocol.  This means that all activity is initiated by the
client side (the network backend)---a server is restricted to
answering requests sent by the client.

The data transferred from the client to the server is comprised of the RPC code
(as a \code{SANE\_WORD}), followed by arguments defined in the {\bf request}
column below.  The format of the server's answer is given in the {\bf reply}
column.

\subsection{\code{\defn{SANE\_NET\_INIT}}}

RPC Code: 0

This RPC establishes a connection to a particular SANE network daemon.
It must be the first call in a SANE network session.  The parameter
and reply arguments for this call are shown in the table below:
\begin{center}
\begin{tabular}{ll}
  {\bf request:} & {\bf reply:} \\
  \code{SANE\_Word version\_code} & \code{SANE\_Word status} \\
  \code{SANE\_String user\_name}   & \code{SANE\_Word version\_code} \\
\end{tabular}
\end{center}
The \code{version\_code} argument in the request is the SANE
version-code of the network backend that is contacting the network
daemon (see Section~\ref{sec:saneversioncode}).  The
``build-revision'' in the version code is used to hold the network
protocol version.  The SANE network daemon receiving such a request
must make sure that the network protocol version corresponds to a
supported version since otherwise the encoding of the network stream
may be incompatible (even though the SANE interface itself may be
compatible).  The \code{user\_name} argument is the name of the user
on whose behalf this call is being performed.  If the network backend
cannot determine a user-name, it passes a \code{NULL} pointer for this
argument.  No trust should be placed in the authenticity of this
user-name.  The intent of this string is to provide more convenience
to the user.  E.g., it could be used as the default-user name in
subsequent authentication calls.

In the reply, \code{status} indicates the completion status.  If the
value is anything other than \code{SANE\_STA\-TUS\_GOOD}, the
remainder of the reply has undefined values.\footnote{The sane network
  daemon should be careful not to leak information in the undefined
  portion of the reply.} The \code{version\_code} argument returns the
SANE version-code that the network daemon supports.  See the comments
in the previous paragraph on the meaning of the build-revision in this
version code.

\subsection{\code{\defn{SANE\_NET\_GET\_DEVICES}}}

RPC Code: 1

This RPC is used to obtain the list of devices accessible by the SANE
daemon.
\begin{center}
\begin{tabular}{ll}
  {\bf request:} & {\bf reply:} \\
  \code{void} & \code{SANE\_Word status} \\
              & \code{SANE\_Device ***device\_list} \\
\end{tabular}
\end{center}
There are no arguments in the request for this call.

In the reply, \code{status} indicates the completion status.  If the
value is anything other than \code{SANE\_STA\-TUS\_GOOD}, the
remainder of the reply has undefined values.  The \code{device\_list}
argument is a pointer to a \code{NULL}-terminated array of
\code{SANE\_Device} pointers.

\subsection{\code{\defn{SANE\_NET\_OPEN}}}

RPC Code: 2

This RPC is used to open a connection to a remote SANE device.
\begin{center}
\begin{tabular}{ll}
  {\bf request:} & {\bf reply:} \\
  \code{SANE\_String device\_name} & \code{SANE\_Word status} \\
                                   & \code{SANE\_Word handle} \\
                                   & \code{SANE\_String resource} \\
\end{tabular}
\end{center}
The \code{device\_name} argument specifies the name of the device to
open.

In the reply, \code{status} indicates the completion status.  If the
value is anything other than \code{SANE\_STA\-TUS\_GOOD}, the
remainder of the reply has undefined values.  The \code{handle}
argument specifies the device handle that uniquely identifies the
connection.  The \code{resource} argument is used to request
authentication.  If it has a non-\code{NULL} value, the network
backend should authenticate the specified resource and then retry this
operation (see Section~\ref{sec:authorization} for details on how to
authorize a resource).

\subsection{\code{\defn{SANE\_NET\_CLOSE}}}

RPC Code: 3

This RPC is used to close a connection to a remote SANE device.
\begin{center}
\begin{tabular}{ll}
  {\bf request:} & {\bf reply:} \\
  \code{SANE\_Word handle} & \code{SANE\_Word dummy} \\
\end{tabular}
\end{center}
The \code{handle} argument identifies the connection that should be
closed.

In the reply, the \code{dummy} argument is unused.  Its purpose is to
ensure proper synchronization (without it, a net client would not be
able to determine when the RPC has completed).

\subsection{\code{\defn{SANE\_NET\_GET\_OPTION\_DESCRIPTORS}}}

RPC Code: 4

This RPC is used to obtain {\em all\/} the option descriptors for a
remote SANE device.
\begin{center}
\begin{tabular}{ll}
  {\bf request:} & {\bf reply:} \\
  \code{SANE\_Word handle} & \code{Option\_Descriptor\_Array odesc} \\
\end{tabular}
\end{center}
The \code{handle} argument identifies the remote device whose option
descriptors should be obtained.

In the reply, the \code{odesc} argument is used to return the array of
option descriptors.  The option descriptor array has the following
structure:
\begin{quote}\index{Option\_Descriptor\_Array}
\begin{verbatim}
struct Option_Descriptor_Array
  {
    SANE_Word num_options;
    SANE_Option_Descriptor **desc;
  };
\end{verbatim}
\end{quote}


\subsection{\code{\defn{SANE\_NET\_CONTROL\_OPTION}}}

RPC Code: 5

This RPC is used to control (inquire, set, or set to automatic) a
specific option of a remote SANE device.
\begin{center}
\begin{tabular}{ll}
  {\bf request:} & {\bf reply:} \\
  \code{SANE\_Word handle}      & \code{SANE\_Status status} \\
  \code{SANE\_Word option}      & \code{SANE\_Word info} \\
  \code{SANE\_Word action}      & \code{SANE\_Word value\_type} \\
  \code{SANE\_Word value\_type} & \code{SANE\_Word value\_size} \\
  \code{SANE\_Word value\_size} & \code{void *value} \\
  \code{void *value}            & \code{SANE\_String *resource} \\
\end{tabular}
\end{center}
The \code{handle} argument identifies the remote device whose option
should be controlled.  Argument \code{option} is the number (index) of
the option that should be controlled.  Argument \code{action}
specifies what action should be taken (get, set, or set automatic).
Argument \code{value\_type} specifies the type of the option value
(must be one of \code{SANE\_TYPE\_BOOL}, \code{SANE\_TYPE\_INT},
\code{SANE\_TYPE\_FIXED}, \code{SANE\_TYPE\_STR\-ING},
\code{SANE\_TYPE\_BUTTON}).  Argument \code{value\_size} specifies
the size of the option value in number of bytes (see
Section~\ref{sec:valuesize} for the precise meaning of this value).
Finally, argument \code{value} is a pointer to the option value.  It
must be a writeable area that is at least \code{value\_size} bytes
large. (Note that this area must be writable even if the action is to
set the option value.  This is because the backend may not be able to
set the exact option value, in which case the option value is used to
return the next best value that the backend has chosen.)

In the reply, argument \code{resource} is set to the name of the
resource that must be authorized before this call can be retried.  If
this value is non-\code{NULL}, all other arguments have undefined
values (see Section~\ref{sec:authorization} for details on how to
authorize a resource).  Argument \code{status} indicates the
completion status.  If the value is anything other than
\code{SANE\_STA\-TUS\_GOOD}, the remainder of the reply has undefined
values.  The \code{info} argument returns the information on how well
the backend was able to satisfy the request.  For details, see the
description of the corresponding argument in
Section~\ref{sec:control}.  Arguments \code{value\_type} and
\code{value\_size} have the same values as the arguments by the same
name in corresponding request.  The values are repeated here to ensure
that both the request and the reply are self-contained (i.e., they can
be encoded and decoded independently).  Argument \code{value} is holds
the value of the option that has become effective as a result of this
RPC.


\subsection{\code{\defn{SANE\_NET\_GET\_PARAMETERS}}}

RPC Code: 6

This RPC is used to obtain the scan parameters of a remote SANE
device.
\begin{center}
\begin{tabular}{ll}
  {\bf request:} & {\bf reply:} \\
  \code{SANE\_Word handle} & \code{SANE\_Status status} \\
                           & \code{SANE\_Parameters params} \\
\end{tabular}
\end{center}
The \code{handle} argument identifies the connection to the remote
device whose scan parameters should be returned.

In the reply, \code{status} indicates the completion status.  If the
value is anything other than \code{SANE\_STA\-TUS\_GOOD}, the
remainder of the reply has undefined values.  The argument
\code{params} is used to return the scan parameters.

\subsection{\code{\defn{SANE\_NET\_START}}}

RPC Code: 7

This RPC is used to start image acquisition (scanning).
\begin{center}
\begin{tabular}{ll}
  {\bf request:} & {\bf reply:} \\
  \code{SANE\_Word handle} & \code{SANE\_Status status} \\
                           & \code{SANE\_Word port} \\
                           & \code{SANE\_Word byte\_order} \\
                           & \code{SANE\_String resource} \\
\end{tabular}
\end{center}
The \code{handle} argument identifies the connection to the remote
device from which the image should be acquired.

In the reply, argument \code{resource} is set to the name of the
resource that must be authorized before this call can be retried.  If
this value is non-\code{NULL}, all other arguments have undefined
values (see Section~\ref{sec:authorization} for details on how to
authorize a resource).  Argument, \code{status} indicates the
completion status.  If the value is anything other than
\code{SANE\_STA\-TUS\_GOOD}, the remainder of the reply has
undefined values.  The argument \code{port} returns the port number
from which the image data will be available.  To read the image data,
a network client must connect to the remote host at the indicated port
number.  Through this port, the image data is transmitted as a
sequence of data records.  Each record starts with the data length in
bytes.  The data length is transmitted as a sequence of four bytes.
These bytes should be interpreted as an unsigned integer in big-endian
format.  The four length bytes are followed by the number of data
bytes indicated by the length.  Except for byte-order, the data is in
the same format as defined for \code{sane\_read()}.  Since some
records may contain no data at all, a length value of zero is
perfectly valid.  The special length value of \code{0xffffffff} is
used to indicate the end of the data stream.  That is, after receiving
a record length of \code{0xffffffff}, the network client should close
the data connection and stop reading data.

Argument \code{byte\_order} specifies the byte-order of the image
data.  A value of 0x1234 indicates little-endian format, a value of
0x4321 indicates big-endian format.  All other values are presently
undefined and reserved for future enhancements of this protocol.  The
intent is that a network server sends data in its own byte-order and
the client is responsible for adjusting the byte-order, if necessary.
This approach causes no unnecessary overheads in the case where the
server and client byte-order match and puts the extra burden on the
client side when there is a byte-order mismatch.  Putting the burden
on the client-side improves the scalability properties of this
protocol.

\subsection{\code{\defn{SANE\_NET\_CANCEL}}}

RPC Code: 8

This RPC is used to cancel the current operation of a remote SANE
device.
\begin{center}
\begin{tabular}{ll}
  {\bf request:} & {\bf reply:} \\
  \code{SANE\_Word handle} & \code{SANE\_Word dummy} \\
\end{tabular}
\end{center}
The \code{handle} argument identifies the connection whose operation
should be cancelled.

In the reply, the \code{dummy} argument is unused.  Its purpose is to
ensure proper synchronization (without it, a net client would not be
able to determine when the RPC has completed).

\subsection{\code{\defn{SANE\_NET\_AUTHORIZE}}}\label{sec:authorization}
\index{network authorization}

RPC Code: 9

This RPC is used to pass authorization data from the net client to the
net server.
\begin{center}
\begin{tabular}{ll}
  {\bf request:} & {\bf reply:} \\
  \code{SANE\_String resource} & \code{SANE\_Word dummy} \\
  \code{SANE\_String username} & \\
  \code{SANE\_String password} & \\
\end{tabular}
\end{center}
The \code{resource} argument specifies the name of the resource to be
authorized.  This argument should be set to the string returned in the
\code{resource} argument of the RPC reply that required this
authorization call.  The \code{username} and \code{password} are the
name of the user that is accessing the resource and the password for
the specified resource/user pair.

Since the password is not encrypted during network transmission, it is
recommended to use the following extension:

If the server adds the string `\code{\$MD5\$}' to the resource-name followed
by a random string not longer then 128 bytes, the client may answer with the
MD5 digest of the concatenation of the password and the random string. To
differentiate between the MD5 digest and a strange password the client prepends
the MD5 digest with the string `\code{\$MD5\$}'.

In the reply, \code{dummy} is completely unused.  Note that there is
no direct failure indication.  This is unnecessary since a net client
will retry the RPC that resulted in the authorization request until
that call succeeds (or until the request is cancelled). The RPC that resulted
in the authorization request continues after the reply from the client and may
fail with \code{SANE\_STATUS\_ACCESS\_DENIED}.


\subsection{\code{\defn{SANE\_NET\_EXIT}}}

RPC Code: 10

This RPC is used to disconnect a net client from a net server.  There
are no request or reply arguments in this call.  As a result of this
call, the connection between the client and the server that was
established by the \code{SANE\_NET\_INIT} call will be closed.

% Local Variables: 
% mode: latex
% TeX-master: "sane.tex"
% End: 
